MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-04-26 Article.tex
more minor stuff

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulas from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean}, allowing construction of large Boolean models.
While this effort faces many challenges, we believe it is a promising way to study the large-scale complexity of biological systems.
However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficiently scalable simulation tools.

@@

A Boolean signaling model consists of $n$ nodes which are either active or inactive, gaining values 1 or 0 respectively. The \emph{state} of the whole model is represented by a vector $S$ of $n$ Boolean values where $S_i$ represents the value of the $i$-th node. We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$; thus $|\mathcal{S}| = 2^n$.

@@

In the CPU version of MaBoSS, the simulation part is the most computationally demanding part, with up to 80\% of MaBoSS runtime spent by just evaluating the Boolean formulas (the exact number depends on the model). The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issues) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}.

There are multiple ways to optimize the expression trees for GPUs: One may use a linked data structure that is more cache-friendly such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulas as a compact continuous array, or convert it to CNF or DNF (conjunctive or disjunctive normal form) bitmasks that can be easily evaluated by vector instructions. We decided to leave the exact representation choice on the compiler, by encoding the expressions as direct code and using the runtime compilation of GPU code~\cite{nvrtc}. In such an approach, the application reads the model files, writes the formulas as functions in CUDA C++, compiles them using the NVIDIA runtime compiler, and finally runs the simulation --- all without a user intervention.

Using this technique, the Boolean formulas are compiled as functions into a native binary code, which is directly executed by the GPU. As the main advantage, the formulae are encoded in the instructions, preventing unnecessary fetches of the encoded formulas from other memory. At the same time, the compiler may apply a vast spectrum of optimizations on the Boolean formulas, including case analysis and shortcutting, again resulting in faster evaluation.

@@

It is worth noting that the datacenter GPU performs worse than the laptop GPU. Both devices are bottlenecked by the runtime compilation of the Boolean formulas, however, NVIDIA A100 spends on average around $300$ms more on the compilation step. Subtracting the compilation time, A100 is faster for all models. We did not spend time finding the root cause of this discrepancy since the value is negligible and the following benchmarks show that the runtime compilation overhead quickly disappears with increasing model size.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models with sizes ranging from $10$ to $1000$ nodes (x-axis). Each model is simulated with 1 million trajectories. The two panels differ by the inclusion of the runtime GPU code compilation, showing its impact on total run time.}
    \label{fig:synth}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes-compilation-big-NVIDIA Tesla A100 (MaBoSS.GPU).pdf}
    \caption{The time spent in the runtime compilation of the Boolean formulas simulating models with varying numbers of nodes, trajectories, and formula lengths.}
    \label{fig:comp}
\end{figure}

Additionally, Figure~\ref{fig:synth} shows the total runtime of the GPU implementation including the runtime compilation step. Comparing the panels, we observe that the relative runtime compilation overhead quickly disappears with increasing model size. Figure~\ref{fig:comp} shows the results of more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU. We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes), which can be simply explained by the fact that these model properties extend source files which need to be compiled by a linear factor. Notably, as soon as the simulation becomes more complex (e.g., by increasing the number of nodes or simulated trajectories), the compilation time becomes relatively negligible even for models with unrealistically long formulae. This suggests that the runtime compilation is a viable optimization methodology also for much larger models.

@@

Mainly, the parallelization scheme used in MaBoSS.GPU could be enhanced to also parallelize over the evaluation of Boolean formulas. To avoid GPU thread divergence, this would however require a specialized Boolean formula representation, entirely different from the current version of MaBoSS; likely even denying the relative efficiency of the use of runtime compilation. On the other hand, this optimization might decrease the register pressure created by holding the state data, and thus increase the performance on models with thousands of nodes.

>>>>

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
More recently, thanks to advances in databases listing such interactions~\cite{licata2020signor,turei2016omnipath} and to experimental techniques providing information on a bigger number of components, the automatic methods have been designed to infer Boolean formulae from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean}, allowing construction of large Boolean models.
While this effort faces many challenges, we believe it is a promising way to study the large-scale complexity of biological systems.
However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficiently scalable simulation tools.

@@

A Boolean signaling model consists of $n$ nodes that are either active or inactive, gaining values 1 or 0 respectively. The \emph{state} of the whole model is represented by a vector $S$ of $n$ Boolean values where $S_i$ represents the value of the $i$-th node. We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$; thus $|\mathcal{S}| = 2^n$.

@@

In the CPU version of MaBoSS, the simulation part is the most computationally demanding part, with up to 80\% of MaBoSS runtime spent by just evaluating the Boolean formulae (the exact number depends on the model). The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issues) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}.

There are multiple ways to optimize the expression trees for GPUs: One may use a linked data structure that is more cache-friendly such as the van Emde Boas tree layout~\cite{van1975preserving}, or perhaps represent the Boolean formulae as a compact continuous array, or convert it to CNF or DNF (conjunctive or disjunctive normal form) bitmasks that can be easily evaluated by vector instructions. We decided to leave the exact representation choice on the compiler, by encoding the expressions as direct code and using the runtime compilation of GPU code~\cite{nvrtc}. In such an approach, the application reads the model files, writes the formulae as functions in CUDA C++ language, compiles them using the NVIDIA runtime compiler, and finally runs the simulation on GPU --- all without user intervention.

Using this technique, the Boolean formulae are compiled as functions into a native binary code, which is directly executed by the GPU. As the main advantage, the formulae are encoded in the instructions, preventing unnecessary fetches of the encoded formulae from other memory. At the same time, the compiler may apply a vast spectrum of optimizations on the Boolean formulae, including case analysis and shortcutting, again resulting in faster evaluation.

@@

It is worth noting that the datacenter GPU performs worse than the laptop GPU. Both devices are bottlenecked by the runtime compilation of the Boolean formulae, however, NVIDIA A100 spends on average around $300$ms more on the compilation step. Subtracting the compilation time, A100 is faster for all models. We did not spend time finding the root cause of this discrepancy since the value is negligible and the following benchmarks show that the runtime compilation overhead quickly disappears with increasing model size.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models with sizes ranging from $10$ to $1000$ nodes (x-axis). Each model is simulated with 1 million trajectories. The two panels differ by the inclusion of the runtime compilation of the model logic, showing its impact on total run time.}
    \label{fig:synth}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes-compilation-big-NVIDIA Tesla A100 (MaBoSS.GPU).pdf}
    \caption{The ratio of time spent in the runtime compilation of the Boolean formulae in relation to the total runtime, simulating models with varying numbers of nodes, trajectories, and formula lengths.}
    \label{fig:comp}
\end{figure}

Additionally, Figure~\ref{fig:synth} shows the total runtime of the GPU implementation including the runtime compilation step. Comparing the panels, we observe that the relative runtime compilation overhead quickly disappears with increasing model size. Figure~\ref{fig:comp} shows the results of more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU. We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes), which can be simply explained by the fact that these model properties extend source files that need to be compiled by a linear factor. Notably, as soon as the simulation becomes more computationally complex (e.g., by increasing the number of nodes or simulated trajectories), the compilation time becomes relatively negligible even for models with unrealistically long formulae. This suggests that the runtime compilation is a viable optimization methodology also for much larger models.

@@

Mainly, the parallelization scheme used in MaBoSS.GPU could be enhanced to also parallelize over the evaluation of Boolean formulae. To avoid GPU thread divergence, this would however require a specialized Boolean formula representation, entirely different from the current version of MaBoSS; likely even denying the relative efficiency of the use of runtime compilation. On the other hand, this optimization might decrease the register pressure created by holding the state data, and thus increase the performance on models with thousands of nodes.
